SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-06-17 seqbias-paper.tex
supplement: finished

\title{Correcting for bias in high throughput sequencing data}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\footnote{to whom correspondence should be addressed}\hspace{0.5em}
Walter L. Ruzzo\,$^{1,2}$
Xinxia Peng\,$^{3}$
Michael G. Katze$^{3}$
}

@@

We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
remaining cases are plotted in Supplementary Section 3.
%% TODO: what section, for real?

@@

Training our model on more reads leads to more accurate estimation of bias, but
increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring under 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 6.
%% TODO: what actual section?

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, Supplementary Section 4).
%% TODO: which actual section ?
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into ChIP-Seq peak calling algorithms, and potentially improve
accuracy.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see,
Supplementary Section 5).
%% TODO: which actual section ?
Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced on the
Illumina platform display similar patterns of nonuniformity, yet differ in
magnitude, suggesting that batch effects in RNA-Seq remain a legitimate concern
that should not be dismissed without evaluation.

>>>>

\title{Correcting for bias in high-throughput sequencing data}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\footnote{to whom correspondence should be addressed}\hspace{0.5em}
Walter L. Ruzzo\,$^{1,2}$
Xinxia Peng\,$^{3}$
Michael G. Katze$^{3}$
}

@@

We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
remaining cases are plotted in Supplementary Section 4.

@@

Training our model on more reads leads to more accurate estimation of bias, but
increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring under 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 5.

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, Supplementary Section 2).
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into ChIP-Seq peak calling algorithms, and potentially improve
accuracy.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see, Supplementary
Section 3).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
on the Illumina platform display similar patterns of nonuniformity, yet differ
in magnitude, suggesting that batch effects in RNA-Seq remain a legitimate
concern that should not be dismissed without evaluation.

### 2011-06-17 seqbias-supplement.tex
compiling in bioinformatics template
supplement: finished

\documentclass[letterpaper]{article}
\usepackage{qcjstd}
\usepackage{lscape}
\usepackage{comment}
\usepackage{geometry}
\usepackage{graphicx}
%\usepackage{palatcm}

\geometry{left=1in,right=1in, top=1in, bottom=1in}

@@

\qcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}

@@

%\begin{figure}[b]
\begin{center}
\includegraphics[width=0.35\textwidth]{fig/trimmed-freqs.pdf}
\end{center}
%\caption{Nucleotide frequencies for trimmed and untrimmed reads from the
%Mortazavi dataset.}
%\label{fig:trimmed}
%\end{figure}

@@

%\begin{figure}[b]
\begin{center}
\includegraphics[width=0.5\textwidth]{fig/freqs_cao.pdf}
\end{center}
%\caption{Nucleotide frequencies from CHiP-Seq reads.}
%\label{fig:freqs_cao}
%\end{figure}

@@

We did however repeat our analysis using the Kullback-Leibler divergence.  We
trained each method using reads from reads from chromosomes 1--8. The remaining
chromosomes were segmented into 500nt bins, and the KL divergence was sampled
from the 50,000 segments with the highest read counts.

First we plot directly the positional KL divergence, computed identically

@@

Secondly, we summarize these plots, computing one normalized divergence
estimate by summing the KL divergence of positions -20 through 20, and
divided by the unadjusted summed divergence to normalize.

\begin{center}
\includegraphics[width=0.5\textwidth]{fig/cao-kl-delta.pdf}
\end{center}

@@

\section{Additional Plots}

Below we plot the adjusted and unadjusted KL divergence of each sample.

%\begin{figure}[b]
%\begin{center}
\includegraphics[width=\textwidth]{fig/kl-all.pdf}
%\end{center}
%\caption{Positional KL divergence for each dataset and method.}
%\label{fig:kl-all}
%\end{figure}

\bibliographystyle{plain}
\bibliography{seqbias-supplement}

>>>>

\documentclass[letterpaper]{article}
\usepackage{lscape}
\usepackage{comment}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{fontspec}

%\setmainfont{Helvetica}

\geometry{left=1.5in,right=1.5in, top=1in, bottom=1in}

\title{SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data}
\author{Daniel C. Jones}

@@

\maketitle

%\qcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}

@@

\begin{center}
\includegraphics[width=0.35\textwidth]{fig/trimmed-freqs.pdf}
\end{center}

@@

\begin{center}
\includegraphics[width=0.7\textwidth]{fig/freqs_cao.pdf}
\end{center}

@@

We did however repeat our analysis using the Kullback-Leibler divergence.
We used several variations of the method described by Hanset, et. al.,
\cite{Hansen2010}. ``7mer'' estimated the initial heptamer, ``Avg-7mer''
averages the intial two heptamers, and ``4mer'' averages the initial 4mer.

We trained each method using reads from reads from chromosomes 1--8. The
remaining chromosomes were segmented into 500nt bins, and the KL divergence was
sampled from the 50,000 segments with the highest read counts.

Below, we plot directly the positional KL divergence, computed identically

@@

@@

\section{Additional Kullback-Leibler Divergence Plots}

Below we plot the adjusted and unadjusted KL divergence plots, supplementing
those in the results section of the main paper.

\includegraphics[width=\textwidth]{fig/kl-all.pdf}

\section{Runtime-Accuracy Tradeoff}

Training our model on more reads results in more accurate estimation of bias,
but a longer training time. Here we investigate that trade-off by training our
model on progressively more reads from the Mortazavi data set. We use the
median pseuo-coefficient of determination ($R^2$) over exons selected for our
test set. Each point is additionally labeled with the time used to train the
model on one core of a 3Ghz Inteal Xeon processor.

\includegraphics[width=\textwidth]{fig/pois-n.eps}

\bibliographystyle{plain}
\bibliography{seqbias-paper}
